From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2017-11-01 appendixA/appendixA.Rmd
WIP

```{r}
knitr::opts_chunk$set(message=FALSE)
```

```{r}
library(nimble)
library(RcppRoll)
library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
library(adaptivetau)
library(tidyverse)

@@

------------

@@

## Stochastic inflation

Analytical predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.  Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.  Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  Stronger nonlinearities (in particular, a larger second derivative at the equilibrium) can drive larger inflationary effects.

```{r}
p <- list(r = .4, K = 1, sigma = .15, N = 500)
equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2

@@

df <- map_dfr(1:10000, 
              function(rep){
                simulate(cmodel)
                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
              },
              .id = "rep")
```

```{r}
df %>% 
  filter(x > 0) %>% 
  group_by(t) %>%
  summarise(ave = mean(x), sd = sd(x)) %>%
  mutate(equib = equib) %>%
  write_csv("inflation.csv")
```

@@

----

>>>>

```{r}
knitr::opts_chunk$set(message=FALSE, cache = TRUE)
```

```{r libraries}
library(nimble)
library(RcppRoll)
library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
library(adaptivetau)
library(tidyverse)

@@

@@

# Stochastic inflation

Analytical predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.  Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.  Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  As predicted by the system size expansion, the deviation from the deterministic result is of order $N^{-1}$.

Because a substantial number of replicate simulations are required to derive a precise mean state, this example will be significantly slower to run than others.

```{r eval=FALSE}
p <- list(r = .4, K = 1, sigma = .15, N = 500)
equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2

@@

## Perform the simulations
df <- map_dfr(1:10000, 
              function(rep){
                simulate(cmodel)
                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
              },
              .id = "rep")

# Tidy up the result data and write to file, along with analytic prediction for equilibrium
df %>% 
  filter(x > 0) %>% 
  group_by(t) %>%
  summarise(ave = mean(x), sd = sd(x)) %>%
  mutate(equib = equib) %>%
  write_csv("inflation.csv")
```

@@

### 2017-11-01 appendixB/appendixB.Rmd
WIP

Also by the chain rule, we have
\begin{align}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\ud \xi}{\ud t} 
\end{align}
Substituting in \eqref{dxidt}, we find:

@@

\begin{align}
&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
&\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
+&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\end{align}

@@

\begin{align}
\langle n\rangle_s = N\left(1-\frac{e}{c}\right)
\end{align}
while the steady-state fluctuations are given by \eqref{fluctuation dissipation}, \eqref{n var}:
\begin{align}
\sigma_n^2 = N\frac{e}{c}
\end{align}

@@

\begin{align}
	b_n - d_n = \frac{A_x}{\epsilon}
\end{align}

Where $A$ is independent of $\epsilon$.  Similarly we have to scale the sum, which being proportional to the variance must be scaled by
\begin{align}
b_n + d_n = \frac{B_x}{\epsilon^2}
\end{align}

We Taylor expand the step operator $\mathbb E$ in the new variable:
\begin{align}
\mathbb E^k &= 1+ \epsilon k\frac{\partial}{\partial x} + \frac{\epsilon^2 k^2}{2} \frac{\partial^2}{\partial x^2} + \ldots
\end{align}

@@

\begin{align}
\frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) \\
\end{align}

@@

\begin{equation}
\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
\end{equation}

>>>>

Also by the chain rule, we have
\begin{align*}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\ud \xi}{\ud t} 
\end{align*}
Substituting in \eqref{dxidt}, we find:

@@

\begin{align}
&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
&\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
+&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)
\end{align}

@@

\begin{align*}
\langle n\rangle_s = N\left(1-\frac{e}{c}\right)
\end{align*}
while the steady-state fluctuations are given by \eqref{fluctuation dissipation}, \eqref{n var}:
\begin{align*}
\sigma_n^2 = N\frac{e}{c}
\end{align*}

@@

\begin{align*}
	b_n - d_n = \frac{A_x}{\epsilon}
\end{align*}

Where $A$ is independent of $\epsilon$.  Similarly we have to scale the sum, which being proportional to the variance must be scaled by
\begin{align*}
b_n + d_n = \frac{B_x}{\epsilon^2}
\end{align*}

We Taylor expand the step operator $\mathbb E$ in the new variable:
\begin{align*}
\mathbb E^k &= 1+ \epsilon k\frac{\partial}{\partial x} + \frac{\epsilon^2 k^2}{2} \frac{\partial^2}{\partial x^2} + \ldots
\end{align*}

@@

\begin{align*}
\frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) 
\end{align*}

@@

\begin{equation*}
\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
\end{equation*}

### 2017-11-01 paper.Rmd
WIP

Stochasticity is most often thought of as something which obscures
patterns we are trying to infer.  This is a simple consequence of
modeling: as discussed above, variables of interest to the processes
in question are modeled explicitly (e.g. change in population density
as it depends on the current density), everything else is just
noise.  Most statistical inference in ecology is built solely to
deal with this problem: pulling back the veil of noise to reveal
processes of interest.  Unfortunately, given the high levels of
variation and relatively limited data typical in ecological analyses,
this can be a significant challenge, and even as statistical and
computational advances provide more powerful and accessible inference
tools, the task only gets harder.  For instance, few ecological
concepts are more central than the notion of density dependence:
that competition in some form or other should limit a species ability
to grow without bound.  Yet establishing statistical evidence for
density dependence is remarkably difficult, as illustrated by
@Knape2011.  Using time series from the Global Population Dynamics
Database (GPDD) Knape & de Valpine show that once one accounts for
the potential for measurement error in addition to environmental
stochasticity -- an unavoidable source of noise in most ecological
data -- it becomes difficult to find evidence for even the simplest
models of density dependence in the majority of available time
series.  Faced with such uncertainty, it is tempting to throw up
our hands and wait until some prophesied deluge of streaming
micro-sensors stapled to every organism provide enough signal to
beat back the noise. Fortunately, this has not been the path taken.

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event.
Consider Levins' patch model, where $n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.  Figure
1 shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.

>>>>

Stochasticity is most often thought of as something which obscures
patterns we are trying to infer.  This is a simple consequence of
modeling: as discussed above, variables of interest to the processes
in question are modeled explicitly (e.g. change in population density
as it depends on the current density), everything else is just
noise.  Most statistical inference in ecology is built solely to
deal with this problem: pulling back the veil of noise to reveal
processes of interest.  Unfortunately, given the high levels of
variation and relatively limited data typical in ecological analyses,
this can be a significant challenge, and even as statistical and
computational advances provide more powerful and accessible inference
tools, the task only gets harder.

For instance, few ecological
concepts are more central than the notion of density dependence:
that competition in some form or other should limit a species ability
to grow without bound.  Yet establishing statistical evidence for
density dependence is remarkably difficult, as illustrated by
@Knape2011.  Using time series from the Global Population Dynamics
Database (GPDD) Knape & de Valpine show that once one accounts for
the potential for measurement error in addition to environmental
stochasticity -- an unavoidable source of noise in most ecological
data -- it becomes difficult to find evidence for even the simplest
models of density dependence in the majority of available time
series.  Faced with such uncertainty, it is tempting to throw up
our hands and wait until some prophesied deluge of streaming
micro-sensors stapled to every organism provide enough signal to
beat back the noise. Fortunately, this has not been the path taken.

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event. Figure 1
uses this algorithm to compare simulations of the Levins' patch model;
see Appendix A for example code.  In the Levins'  model,
$n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.  Figure
1 shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.
